Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-14 ind_files/99_Appendix_D_examples.tex
Change filenames to be more sensible

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since \eqref{ex_1D_stoch_logistic_det_limit}t only depends on the difference $\lambda-\mu$.
\myfig{0.6}{Media/App_stoch_logistic_same_w_different_t.png}{Comparison of time series (Gillespie simulations) for two different parameter values of the stochastic logistic equation. Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

@@

\myfig{1}{Media/4.1_stoch_logistic_timeseries.png}{Comparison of a single realization of the exact birth-death process \eqref{ex_1D_stoch_logistic_BD_eqns}, the deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit}, the non-linear Fokker-Planck equation \eqref{ex_1D_stoch_logistic_full_SDE}, and the weak noise approximation \eqref{ex_1D_stoch_logistic_WNA} for \textbf{(A)} $K = 500$, \textbf{(B)} $K = 1000$, and \textbf{(C)} $K = 10000$. $\lambda = 2, \mu = 1$ for all thee cases.}{fig_1D_stoch_logistic_timeseries}
The deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit} has two fixed points, one at $x=0$ (extinction) and one at $x=1$ (corresponding to a population size of $n=K$). For $r > 0$, $x=0$ is unstable and $x=1$ is a global attractor, meaning in the deterministic limit, when $r > 0$, all populations end up at $x=1$ given enough time.\myfig{0.85}{Media/4.2_stoch_logistic_distributions.png}{Comparison of the steady-state densities given by \eqref{ex_1D_stoch_logistic_BD_eqns}, \eqref{ex_1D_stoch_logistic_full_SDE}, and \eqref{ex_1D_stoch_logistic_WNA} for \textbf{(A)} $K = 500$, \textbf{(B)} $K = 1000$, and \textbf{(C)} $K = 10000$. $\lambda = 2, \mu = 1$ for all thee cases. Each curve was obtained using $1000$ independent realizations.}{fig_1D_stoch_logistic_densities}

@@

In Chapter \ref{chap_infD_processes}, we used various approximations to arrive at the linear functional Fokker-Planck equation
\begin{equation}
\label{functional_WNE_chap_4}
    \frac{\partial P}{\partial t}(\zeta,t) = \int\limits_{\mathcal{T}}\left(-\frac{\delta}{\delta \zeta(x)}\left\{\mathcal{D}_{\zeta}[\mathcal{A}^{-}](x)P(\zeta,t)\right\}+\frac{1}{2}\mathcal{A}^{+}(x|\psi)\frac{\delta^2}{\delta\zeta(x)^2}\{P(\zeta,t)\}\right)dx
\end{equation}
for describing stochastic fluctuations $\zeta$ from the deterministic solution obtained by solving \eqref{deterministic_traj}. Our goal is now to find a method to effectively detect and describe evolutionary branches (modes in trait space, corresponding to individual morphs) for this process. Following the methods used by Tim Rogers and colleagues for various special cases~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015}, we will do this in a general manner by measuring the autocorrelation of the distribution of the population over trait space, a task made easier by moving to Fourier space.
\myfig{0.8}{Media/4.3_fourier_fig.png}{\textbf{Schematic description of Fourier analysis}. A function $\phi(x)$ (shown in red) over the trait space can be decomposed as the sum of infinitely many Fourier modes (shown in blue) $\phi_k$. In the Fourier dual space, we can look at the peaks of each of these Fourier modes: The magnitude of $\phi_k$ tells us how much it contributes to the actual function of interest $\phi$.}{fig_Fourier}
Specifically, a convenient theorem due to Weiner and Khinchin relates the autocorrelation of a probability distribution to its power spectral density via Fourier transformation. This has been extensively used in spatial ecology, and we too will make use of it here by carrying out a basis expansion of our functions in the Fourier basis $\{e^{ikx}\}_{k\in\mathbb{Z}}$. If $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ is a translation-invariant\footnote{This is horrible nomenclature by the mathematicians. Though `invariant' is the conventional name for this concept, the intended meaning is not really invariant but `equivariant'. Formally, let $\mathcal{F}$ be a suitable function space of real valued functions. For any $c \in \mathbb{R}$, let $T_c: \mathcal{F} \to \mathcal{F}$ be the translation operator on this space, defined by $T_c[f(x)] = f(x+c)$. An operator $L: \mathcal{F} \to \mathcal{F}$ is said to be translation-invariant if it commutes with $T_c$ for every $c \in \mathbb{R}$, \emph{i.e.} $T_c[L[f]] = L[T_c[f]] \ \forall \ f \in \mathcal{F} \ \forall \ c \in \mathbb{R}$. This is not as restrictive as it initially sounds. For example, both the Laplacian operator and the convolution operator are linear and translation invariant} linear operator, then $\exp(ikx)$ acts as an eigenfunction, significantly simplifying the calculations. We therefore assume that $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ takes the form:
\begin{equation*}
 \mathcal{D}_{\zeta}[\mathcal{A}^-](x,t) = L[\zeta(x,t)]   
\end{equation*}
for a translation-invariant linear operator $L$ that only depends on $x$ and $t$. The presence of phenotypic clustering and polymorphisms can be analyzed by examining the power spectrum of $\Tilde{P}_{0}(\zeta,s)$ over the trait space.

@@

We can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}
Using this in equation \eqref{spatial_covariance_zeta},
\cite{rogers_demographic_2012} (and later~\cite{rogers_modes_2015}) have shown that the contribution of demographic stochasticity can lead to inhibition of branching, and thus, while the population undergoes infinitely many branching events in the infinite population prediction, this does not happen for finite populations.
\myfig{0.8}{Media/2.1_dem_stoch_effects.png}{\textbf{Effect of population size on evolutionary branching}. Two different realizations of the system \eqref{Rogers_logistic_BD} with $n(x) = \exp(-x^2/\sigma_{K}^2)$. Simulation parameters are $\sigma_{K} = 1.9, \sigma_{\alpha} = 0.7, \sigma^2_{m} = 0.05$ for \textbf{top:} $K = 1000$ and \textbf{bottom:} $K = 10000$. Each point represents an individual. Note that the model on top remains monomorphic whereas the model on the bottom exhibits evolutionary branching, where an initially monomorphic population evolves to become dimorphic.}{fig_dem_stoch}
An alternative `moment-based' method that avoids moving to Fourier space has also been used to study this phenomenon of evolutionary branching and clustering in finite populations~\citep{wakano_evolutionary_2013,debarre_evolutionary_2016}. These studies use the equation we derived in section \ref{sec_fun_theorems_var} for the variance of the trait in the population and compute the conditions required for the variance to `explode' (Equation A.23 in~\cite{debarre_evolutionary_2016} is exactly equivalent to equation \eqref{nD_stochastic_Price_variance} for their choice of functional forms upon converting their change in variance to an infinitesimal rate of change \emph{i.e.} a derivative). The method itself is relatively straightforward in principle (complications arise if the particular models are complicated) and I therefore do not explore it further in this thesis, but the broad results of such moment-based approaches is in agreement with the predictions made from the spectral methods employed in~\cite{rogers_demographic_2012} and~\cite{rogers_modes_2015}.

>>>>

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since \eqref{ex_1D_stoch_logistic_det_limit}t only depends on the difference $\lambda-\mu$.
\myfig{0.6}{figures/App_stoch_logistic_same_w_different_t.png}{Comparison of time series (Gillespie simulations) for two different parameter values of the stochastic logistic equation. Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

@@

\myfig{1}{figures/App_stoch_logistic_timeseries.png}{Comparison of a single realization of the exact birth-death process \eqref{ex_1D_stoch_logistic_BD_eqns}, the deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit}, the non-linear Fokker-Planck equation \eqref{ex_1D_stoch_logistic_full_SDE}, and the weak noise approximation \eqref{ex_1D_stoch_logistic_WNA} for \textbf{(A)} $K = 500$, \textbf{(B)} $K = 1000$, and \textbf{(C)} $K = 10000$. $\lambda = 2, \mu = 1$ for all thee cases.}{fig_1D_stoch_logistic_timeseries}
The deterministic trajectory \eqref{ex_1D_stoch_logistic_det_limit} has two fixed points, one at $x=0$ (extinction) and one at $x=1$ (corresponding to a population size of $n=K$). For $r > 0$, $x=0$ is unstable and $x=1$ is a global attractor, meaning in the deterministic limit, when $r > 0$, all populations end up at $x=1$ given enough time.\myfig{0.85}{figures/App_stoch_logistic_distributions.png}{Comparison of the steady-state densities given by \eqref{ex_1D_stoch_logistic_BD_eqns}, \eqref{ex_1D_stoch_logistic_full_SDE}, and \eqref{ex_1D_stoch_logistic_WNA} for \textbf{(A)} $K = 500$, \textbf{(B)} $K = 1000$, and \textbf{(C)} $K = 10000$. $\lambda = 2, \mu = 1$ for all thee cases. Each curve was obtained using $1000$ independent realizations.}{fig_1D_stoch_logistic_densities}

@@

In Chapter \ref{chap_infD_processes}, we used various approximations to arrive at the linear functional Fokker-Planck equation
\begin{equation}
\label{functional_WNE_chap_4}
    \frac{\partial P}{\partial t}(\zeta,t) = \int\limits_{\mathcal{T}}\left(-\frac{\delta}{\delta \zeta(x)}\left\{\mathcal{D}_{\zeta}[\mathcal{A}^{-}](x)P(\zeta,t)\right\}+\frac{1}{2}\mathcal{A}^{+}(x|\psi)\frac{\delta^2}{\delta\zeta(x)^2}\{P(\zeta,t)\}\right)dx
\end{equation}
for describing stochastic fluctuations $\zeta$ from the deterministic solution obtained by solving \eqref{deterministic_traj}. Our goal is now to find a method to effectively detect and describe evolutionary branches (modes in trait space, corresponding to individual morphs) for this process. Following the methods used by Tim Rogers and colleagues for various special cases~\citep{rogers_demographic_2012, rogers_spontaneous_2012, rogers_modes_2015}, we will do this in a general manner by measuring the autocorrelation of the distribution of the population over trait space, a task made easier by moving to Fourier space.
\myfig{0.8}{figures/App_fourier_fig.png}{\textbf{Schematic description of Fourier analysis}. A function $\phi(x)$ (shown in red) over the trait space can be decomposed as the sum of infinitely many Fourier modes (shown in blue) $\phi_k$. In the Fourier dual space, we can look at the peaks of each of these Fourier modes: The magnitude of $\phi_k$ tells us how much it contributes to the actual function of interest $\phi$.}{fig_Fourier}
Specifically, a convenient theorem due to Weiner and Khinchin relates the autocorrelation of a probability distribution to its power spectral density via Fourier transformation. This has been extensively used in spatial ecology, and we too will make use of it here by carrying out a basis expansion of our functions in the Fourier basis $\{e^{ikx}\}_{k\in\mathbb{Z}}$. If $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ is a translation-invariant\footnote{This is horrible nomenclature by the mathematicians. Though `invariant' is the conventional name for this concept, the intended meaning is not really invariant but `equivariant'. Formally, let $\mathcal{F}$ be a suitable function space of real valued functions. For any $c \in \mathbb{R}$, let $T_c: \mathcal{F} \to \mathcal{F}$ be the translation operator on this space, defined by $T_c[f(x)] = f(x+c)$. An operator $L: \mathcal{F} \to \mathcal{F}$ is said to be translation-invariant if it commutes with $T_c$ for every $c \in \mathbb{R}$, \emph{i.e.} $T_c[L[f]] = L[T_c[f]] \ \forall \ f \in \mathcal{F} \ \forall \ c \in \mathbb{R}$. This is not as restrictive as it initially sounds. For example, both the Laplacian operator and the convolution operator are linear and translation invariant} linear operator, then $\exp(ikx)$ acts as an eigenfunction, significantly simplifying the calculations. We therefore assume that $\mathcal{D}_{\zeta}[\mathcal{A}^{-}]$ takes the form:
\begin{equation*}
 \mathcal{D}_{\zeta}[\mathcal{A}^-](x,t) = L[\zeta(x,t)]   
\end{equation*}
for a translation-invariant linear operator $L$ that only depends on $x$ and $t$. The presence of phenotypic clustering and polymorphisms can be analyzed by examining the power spectrum of $\Tilde{P}_{0}(\zeta,s)$ over the trait space.

@@

We can also calculate $\mathcal{D}_{\zeta}[\mathcal{A}^-]$ as
\begin{align*}
\mathcal{D}_{\zeta}[\mathcal{A}^-] &= \frac{d}{d\epsilon}\left( r\int\limits_{\mathcal{T}} m(x,y)(\psi(y)+\epsilon\zeta(y))dy - \frac{\psi(x)+\epsilon\zeta(x)}{n(x)}\int\limits_{\mathcal{T}}\alpha(x,y)(\psi(y)+\epsilon\zeta(y))dy\right) \biggl{|}_{\epsilon = 0}\\
&= r\int\limits_{\mathcal{T}}m(x,y)\zeta(y)dy - \frac{1}{n(x)}\left(\psi(x)\int\limits_{\mathcal{T}}\alpha(x,y)\zeta(y)dy + \zeta(x)\int\limits_{\mathcal{T}}\alpha(x,y)\psi(y)dy\right)
\end{align*}
Using this in equation \eqref{spatial_covariance_zeta},
\cite{rogers_demographic_2012} (and later~\cite{rogers_modes_2015}) have shown that the contribution of demographic stochasticity can lead to inhibition of branching, and thus, while the population undergoes infinitely many branching events in the infinite population prediction, this does not happen for finite populations.
\myfig{0.8}{figures/App_dem_stoch_effects.png}{\textbf{Effect of population size on evolutionary branching}. Two different realizations of the system \eqref{Rogers_logistic_BD} with $n(x) = \exp(-x^2/\sigma_{K}^2)$. Simulation parameters are $\sigma_{K} = 1.9, \sigma_{\alpha} = 0.7, \sigma^2_{m} = 0.05$ for \textbf{top:} $K = 1000$ and \textbf{bottom:} $K = 10000$. Each point represents an individual. Note that the model on top remains monomorphic whereas the model on the bottom exhibits evolutionary branching, where an initially monomorphic population evolves to become dimorphic.}{fig_dem_stoch}
An alternative `moment-based' method that avoids moving to Fourier space has also been used to study this phenomenon of evolutionary branching and clustering in finite populations~\citep{wakano_evolutionary_2013,debarre_evolutionary_2016}. These studies use the equation we derived in section \ref{sec_fun_theorems_var} for the variance of the trait in the population and compute the conditions required for the variance to `explode' (Equation A.23 in~\cite{debarre_evolutionary_2016} is exactly equivalent to equation \eqref{nD_stochastic_Price_variance} for their choice of functional forms upon converting their change in variance to an infinitesimal rate of change \emph{i.e.} a derivative). The method itself is relatively straightforward in principle (complications arise if the particular models are complicated) and I therefore do not explore it further in this thesis, but the broad results of such moment-based approaches is in agreement with the predictions made from the spectral methods employed in~\cite{rogers_demographic_2012} and~\cite{rogers_modes_2015}.
